---
title: "UTK MSBA Datathon"
author: "Author"
date: "2022-11-15"
output: html_document
---

The data includes median housing price (Zillow), median rent (Zillow), household income (Census), and household count (Census) by zip code for the past few years. The data also includes interest rate for 30-year mortgage loans in the past years.

```{r}
library(tidyverse)
library(stringr)
```

# Interest Rate

```{r}
Mortgage = read.csv('https://drive.google.com/uc?id=1lokCssSi25y6MzXBtry9QMOTfkZg0btW') %>%
  `colnames<-`(c('Year', 'Rate')) %>%
  mutate(Rate=as.numeric(str_replace(Rate, '%', '')))
#Mortgage

```

```{r}
plot(Mortgage, type = 'l')
```

# **Zip: Households and Income**

```{r}
Zip = read_csv('https://drive.google.com/uc?id=12wY5XI1KfV9ffmUDjO2utTrSa5Ksyus8') %>%
  `colnames<-`(c('Year', 'Zip_Code', 'Households', 'Median_Income_USD')) %>%
  filter(Households != 'null' | Median_Income_USD != 'null') %>%
  mutate(Households = as.numeric(Households))
#Zip
```

```{r}
Zip <- Zip %>%
  filter(!str_detect(Median_Income_USD, '\\+'))  ## REMOVED median income with a +
```

```{r}
Zip <- Zip %>%
  filter(!str_detect(Median_Income_USD, '\\-')) ## REMOVED median income with a -

Zip <- Zip %>%
  filter(!str_detect(Median_Income_USD, '(X)')) ## REMOVED median income with a (X)

Zip$Median_Income_USD <- as.integer(Zip$Median_Income_USD) ## CHANGED median income to integer


```

# HOME VALUES INDEX

Zillow Home Value Index (ZHVI): A smoothed, seasonally adjusted measure of the typical home value and market changes across a given region and housing type. It reflects the typical value for homes in the 35th to 65th percentile range. The raw version of that mid-tier ZHVI time series is also available.

Zillow publishes top-tier ZHVI (\$, typical value for homes within the 65th to 95th percentile range for a given region) and bottom-tier ZHVI (\$, typical value for homes that fall within the 5th to 35th percentile range for a given region).

Zillow also publishes ZHVI for all single-family residences (\$, typical value for all single-family homes in a given region), for condo/coops (\$), for all homes with 1, 2, 3, 4 and 5+ bedrooms (\$), and the ZHVI per square foot (\$, typical value of all homes per square foot calculated by taking the estimated home value for each home in a given region and dividing it by the home's square footage). Check out this overview of ZHVI and a deep-dive into its methodology.

Here's a handy ZHVI User Guide for information about properly citing and making calculations with this metric.

```{r}
VI = read.csv('https://drive.google.com/uc?id=1ZToY3SzTWwg0gbKj_H9o7h6myt_dKFIu', row.names ='RegionName') %>% 
   as_tibble()
#VI
```

```{r}
VIMonth = select(VI, contains('.'))
#VIMonth
```

```{r}
VITime = cbind(RegionName=rownames(VIMonth), stack(VIMonth)) %>%
  mutate(
    Year = substr(ind, 2, 5),
    Month = substr(ind, 7, 8),
  )
#VITime

```

```{r}
VIRegion = select(VI, !contains('.'))
#VIRegion
```


```{r}
VIRegion <- VIRegion %>% mutate(Bind_Col = 1:nrow(VIRegion)) ## Made RegionName to join later.
VIRegion$Bind_Col <- as.integer(VIRegion$Bind_Col) # CHANGED Bind_Col to integer
VITime$RegionName <- as.integer(VITime$RegionName) # CHANGED RegionName to integer
VITime$Month <- NULL ## Removed Month column as it caused conflicts. 
VIRegion_VITime_Joined <- full_join(VIRegion,VITime, by = c("Bind_Col" = "RegionName")) # JOINING ALL DATA BY RegionName

VIRegion_VITime_Joined <- VIRegion_VITime_Joined %>% mutate(Month = substr(ind, 7, 8)) ## ADDED Month Columns back to be used for a join later

colnames(VIRegion_VITime_Joined)[10] <- "ZHVI" # CHANGED the median house price from index to ZHVI. THIS IS THE average of the 35th - 65th percentile for housing price.

VIRegion_VITime_Joined$Year <- as.integer(VIRegion_VITime_Joined$Year) ## CHANGED Year to integer
VIRegion_VITime_Joined <- full_join(VIRegion_VITime_Joined, Mortgage, by = "Year") ## JOINED ALL DATA from mortgage to VIRegion_VITime
###########################################
#VIRegion_VITime_Joined      This dataset is all the housing index stuff we need.
############################################
```


```{r}
## CLEANING UP
rm(VI)
rm(VIMonth)
rm(VITime)
rm(VIRegion)
## Only the Data we need for now
```

# RENT INDEX

Zillow Observed Rent Index (ZORI): A smoothed measure of the typical observed market rate rent across a given region. ZORI is a repeat-rent index that is weighted to the rental housing stock to ensure representativeness across the entire market, not just those homes currently listed for-rent. The index is dollar-denominated by computing the mean of listed rents that fall into the 40th to 60th percentile range for all homes and apartments in a given region, which is once again weighted to reflect the rental housing stock. Details available in ZORI methodology.

```{r}
RI = read.csv('https://drive.google.com/uc?id=1MTnf6k39PShycq66yRANKqFZW5BmJfME')
#RI
```

```{r}
RIMonth = select(RI, contains('.'))
#RIMonth
```

```{r}
RITime = cbind(RegionName=rownames(RIMonth), stack(RIMonth)) %>%
  mutate(
    Year = substr(ind, 2, 5),
    Month = substr(ind, 7, 8),
  )
#RITime
```

```{r}
RIRegion = select(RI, !contains('.'))
#RIRegion

ZIP_CODE_RI <- RIRegion %>% select(RegionID,RegionName)

VIRegion_VITime_Joined_RI_ZIP <- inner_join(VIRegion_VITime_Joined, ZIP_CODE_RI, by = "RegionID")
VIRegion_VITime_Joined_RI_ZIP$Year <- as.integer(VIRegion_VITime_Joined_RI_ZIP$Year) 
HOUSE_WITH_EVERYTHING <- inner_join(VIRegion_VITime_Joined_RI_ZIP, Zip, by = c("Year","RegionName" = "Zip_Code")) 

HOUSE_WITH_EVERYTHING <- HOUSE_WITH_EVERYTHING %>% filter(Year < 2020)
colnames(HOUSE_WITH_EVERYTHING)[10] <- "ZHVI"
```


```{r}
RIRegion <- RIRegion %>% mutate(Bind_Col = 1:nrow(RIRegion)) ## Made RegionName to join later.
RITime$Month <- NULL ## Removed Month column as it caused conflicts. CAN BE ADDED LATER IF NEEDED
RITime$RegionName <- as.integer(RITime$RegionName)

RIRegion_RITime_Joined <- full_join(RIRegion,RITime, by = c("Bind_Col" = "RegionName")) # JOINING ALL DATA BY RegionName 

RIRegion_RITime_Joined <- RIRegion_RITime_Joined %>% mutate(Month = substr(ind, 7, 8)) # ADDING Month back for the join later

colnames(RIRegion_RITime_Joined)[11] <- "ZORI" # CHANGED the median rent price from index to ZORI

RIRegion_RITime_Joined$Year <- as.integer(RIRegion_RITime_Joined$Year) ## CHANGED Year to integer

RIRegion_RITime_Joined_2015_Mortage_Join <- full_join(RIRegion_RITime_Joined, Mortgage, by = "Year") ## JOINED ALL DATA from mortgage to RIRegion_RITime_Joined

###
# RIRegion_RITime_Joined ## Is all of the Renter Index information Together. 
###
```


```{r}
FULL_DATASET <- full_join(VIRegion_VITime_Joined,RIRegion_RITime_Joined_2015_Mortage_Join, by = c("RegionID","Metro","ind","Rate","StateName","Year","Month","CountyName","State","City","SizeRank","RegionType","Bind_Col"))## INNER JOINING by RegionName, Month, and Year in order to match everything correctly. The suffix is there to understand which dataset you are looking at. _VI refers to the housing value index. _RI refers to the Renter Index. 

#There is a lot of missing data for the index's as one index will have data but the other wont for that time of the year and month. Should be okay as we are mainly looking at houses.

FULL_DATASET <- FULL_DATASET %>% filter(Year <= 2021) ## Removed dates that go beyond the 30 - year mortage rate.

FULL_DATASET <- full_join(FULL_DATASET,Zip, by = c("RegionName" = "Zip_Code", "Year"))

colnames(FULL_DATASET)[15] <- "Zip_Code"
```

```{r}
#CLEANING UP
rm(RI)
rm(RIMonth)
rm(RIRegion)
rm(RITime)
rm(zip_2015)
rm(RIRegion_RITime_Joined)
rm(VIRegion_VITime_Joined_2015_Mortage_Join)
rm(Mortgage)
rm(Zip)
rm(RIRegion_RITime_Joined_2015_Mortage_Join)
rm(VIRegion_VITime_Joined)
```

```{r}
TEST <- FULL_DATASET %>%
  group_by(Year) %>%
  summarise(mean_ZHVI = mean(ZHVI, na.rm = T),
            mean_ZORI = mean(ZORI, na.rm = T))

ggplot(TEST, aes(Year, mean_ZHVI)) + geom_point() + geom_line()

# This is just for testing purposes.

```

```{r}
PMT <- function(PV, i, n) {
(PV*i*(1+i)^n) / ((1+i)^n-1)
}

PMT(175000, 0.04125/12, 30*12)

PV <- function(PMT, i, n) {
(PMT*((1+i)^n-1)) / (i*(1+i)^n)
}

PV(948.61, 0.03/12, 30*12)

HOUSE_WITH_EVERYTHING$MonthlyPmt <- c()
for (i in 1:nrow(HOUSE_WITH_EVERYTHING)) {
  HOUSE_WITH_EVERYTHING$MonthlyPmt[i] = PMT(HOUSE_WITH_EVERYTHING$ZHVI[i], (HOUSE_WITH_EVERYTHING$Rate[i]/100)/12, 30*12)
}
HOUSE_WITH_EVERYTHING <- HOUSE_WITH_EVERYTHING %>% 
                          mutate(MonthlyPmt = PMT(ZHVI,(Rate/100)/12,30*12),
                                 QualifyingIncome = MonthlyPmt * 4 * 12,
                                 AffordabilityIndex = (Median_Income_USD/QualifyingIncome) * 100)
summary(HOUSE_WITH_EVERYTHING, na.rm = T)

hist(HOUSE_WITH_EVERYTHING$AffordabilityIndex)
mean(HOUSE_WITH_EVERYTHING$AffordabilityIndex < 100, na.rm = T)
mean(HOUSE_WITH_EVERYTHING$AffordabilityIndex >= 100, na.rm = T)

```   

