---
title: "UTK MSBA Datathon"
author: "Author"
date: "2022-11-15"
output: html_document
---

The data includes median housing price (Zillow), median rent (Zillow), household income (Census), and household count (Census) by zip code for the past few years. The data also includes interest rate for 30-year mortgage loans in the past years.

```{r}
library(tidyverse)
library(stringr)
```

# Interest Rate

```{r}
# add mortgage data, remove % from values
Mortgage = read.csv('https://drive.google.com/uc?id=1lokCssSi25y6MzXBtry9QMOTfkZg0btW') %>%
  `colnames<-`(c('Year', 'Rate')) %>%
  mutate(Rate=as.numeric(str_replace(Rate, '%', '')))
Mortgage
```

```{r}
# Time series plot of mortgage rates from 1970s - ~2022
plot(Mortgage, type = 'l')
```

# **Zip: Hourseholds and Income**

```{r}
# Load in zip code data, remove null households and null incomes, convert households to numeric
Zip = read_csv('https://drive.google.com/uc?id=12wY5XI1KfV9ffmUDjO2utTrSa5Ksyus8') %>%
  `colnames<-`(c('Year', 'Zip_Code', 'Households', 'Median_Income_USD')) %>%
  filter(Households != 'null' | Median_Income_USD != 'null') %>%
  mutate(Households = as.numeric(Households))
Zip
```

```{r}
# There are 136 instances with a median income > 250,000+
Zip %>%
  filter(str_detect(Median_Income_USD, '\\+')) %>%
  group_by(Median_Income_USD) %>%
  summarise(n())
```

```{r}
# 122 instances of median income 2,500 or less
Zip %>%
  filter(str_detect(Median_Income_USD, '\\-')) %>%
  group_by(Median_Income_USD) %>%
  summarise(n())
```

# HOME VALUES INDEX

Zillow Home Value Index (ZHVI): A smoothed, seasonally adjusted measure of the typical home value and market changes across a given region and housing type. It reflects the typical value for homes in the 35th to 65th percentile range. The raw version of that mid-tier ZHVI time series is also available.

Zillow publishes top-tier ZHVI (\$, typical value for homes within the 65th to 95th percentile range for a given region) and bottom-tier ZHVI (\$, typical value for homes that fall within the 5th to 35th percentile range for a given region).

Zillow also publishes ZHVI for all single-family residences (\$, typical value for all single-family homes in a given region), for condo/coops (\$), for all homes with 1, 2, 3, 4 and 5+ bedrooms (\$), and the ZHVI per square foot (\$, typical value of all homes per square foot calculated by taking the estimated home value for each home in a given region and dividing it by the home's square footage). Check out this overview of ZHVI and a deep-dive into its methodology.

Here's a handy ZHVI User Guide for information about properly citing and making calculations with this metric.

```{r}
# Load in housing zillow data
VI = read.csv('https://drive.google.com/uc?id=1ZToY3SzTWwg0gbKj_H9o7h6myt_dKFIu', row.names ='RegionName') %>% 
   as_tibble()
VI
```

```{r}
# These columns represent the median housing prices according to zillow (I think)
VIMonth = select(VI, contains('.'))
VIMonth
```

```{r}
# Add variable RegionName for row # (seems useless). Seperate year and month from the dates
VITime = cbind(RegionName=rownames(VIMonth), stack(VIMonth)) %>%
  mutate(
    Year = substr(ind, 2, 5),
    Month = substr(ind, 7, 8),
  )
VITime
```

```{r}
# select columns without a .
VIRegion = select(VI, !contains('.'))
VIRegion
```

# RENT INDEX

Zillow Observed Rent Index (ZORI): A smoothed measure of the typical observed market rate rent across a given region. ZORI is a repeat-rent index that is weighted to the rental housing stock to ensure representativeness across the entire market, not just those homes currently listed for-rent. The index is dollar-denominated by computing the mean of listed rents that fall into the 40th to 60th percentile range for all homes and apartments in a given region, which is once again weighted to reflect the rental housing stock. Details available in ZORI methodology.

```{r}
# Add rent data
RI = read.csv('https://drive.google.com/uc?id=1MTnf6k39PShycq66yRANKqFZW5BmJfME', row.names='RegionName')
RI
```

```{r}
# select median monthly rent columns
RIMonth = select(RI, contains('.'))
RIMonth
```

```{r}
# Add column for row numbers, seperate year and month
RITime = cbind(RegionName=rownames(RIMonth), stack(RIMonth)) %>%
  mutate(
    Year = substr(ind, 2, 5),
    Month = substr(ind, 7, 8),
  )
RITime
```

```{r}
# select all other columns
RIRegion = select(RI, !contains('.'))
RIRegion
```



```{r}
# Join mortgage and zip code data, filter out 0 households
Mortgage_Zip <- inner_join(Mortgage,Zip, by = c("Year"))
Mortgage_Zip_VI <- inner_join(Mortgage_Zip,VI, by = c("Zip_Code" = "RegionID"))
Mortgage_Zip_VI_Zero_Removed <- Mortgage_Zip_VI %>% filter(!Households == 0)
```

```{r}
#Convert year to factor
RITime$Year <- as.factor(RITime$Year)
Mortgage$Year <- as.factor(Mortgage$Year)

# Join rent index with mortgage data
RITime_Mortgage  <- inner_join(RITime,Mortgage, by = c("Year"))

# convert variables to factor
RITime_Mortgage$RegionName <- as.factor(RITime_Mortgage$RegionName)
RIRegion$RegionID <- as.factor(RIRegion$RegionID)

# join rent/mortgage data
RITime_Mortgage_Region <- inner_join(RITime_Mortgage,RIRegion, by = c("RegionName" = "RegionID"))

# convert variables to factor
RITime_Mortgage_Region$RegionName <- as.factor(RITime_Mortgage_Region$RegionName)
Zip$Year <- as.factor(Zip$Year)
Zip$Zip_Code <- as.factor(Zip$Zip_Code)

# Join rent/mortgage/region data with zip code data
RITime_Mortage_Region_Census <- inner_join(RITime_Mortgage_Region,Zip, by = c("RegionName" = "Zip_Code", "Year" = "Year"))

# Average the median rent and median income aggregated by year
RENTAL <- RITime_Mortage_Region_Census %>%
  group_by(Year) %>%
  drop_na(values) %>%
  summarise(mean_of_median_rent = mean(values),
            mean_of_median_income = mean(as.numeric(Median_Income_USD)))

#Mortage_Rate <- RITime_Mortage_Region_Census %>%
  
```

```{r}
library(ggplot2)
names(Zip)

#Convert median income to numeric
Zip$Median_Income_USD <- as.numeric(Zip$Median_Income_USD)

# Find the maximum median income for a zip-code
Zip[which.max(Zip$Median_Income_USD), ]

# Yearly median income across all zip-codes
YearlyIncome <- Zip %>% 
  group_by(Year) %>% 
  summarise(
    medIncome = median(Median_Income_USD, na.rm = TRUE)
  )

ggplot(YearlyIncome, mapping = aes(x = Year, y = medIncome)) + geom_line()

# Convert median income to numeric
Mortgage_Zip_VI$Median_Income_USD <- as.numeric(Mortgage_Zip_VI$Median_Income_USD)

# Median income, maximum yearly and monthly mortgage payments aggregated by zip code
MyShit <- Mortgage_Zip_VI %>% 
  filter(Year %in% c(2020)) %>% 
  group_by(Zip_Code) %>% 
  summarise(
    medIncome = Median_Income_USD,
    maxYrPmt = Median_Income_USD * 0.25,
    maxMthPmt = (Median_Income_USD * 0.25)/12
  )

# remove NA incomes and round max payments to 2 decimal places
MyShit <- MyShit[!is.na(MyShit$medIncome), ]
MyShit$maxYrPmt <- round(MyShit$maxYrPmt, 2)
MyShit$maxMthPmt <- round(MyShit$maxMthPmt, 2)

# Median income, max yrly/mthly pmnt, and max household price by zip code (need compounding logic for hh price)
MyShit <- MyShit %>% 
  group_by(Zip_Code) %>% 
  summarise(
    medIncome = medIncome,
    maxYrPmt = maxYrPmt,
    maxMthPmt = maxMthPmt,
    maxHHPrice = maxMthPmt * 30 + (maxMthPmt * 30 * 0.01) # NEEDS TO COMPOUND
  )

# Remove NA incomes
MyShit <- MyShit[!is.na(MyShit$medIncome), ]

# Histogram of max household prices across all zip-codes
ggplot(MyShit, mapping = aes(x = maxHHPrice)) + geom_histogram(binwidth = 2500)
```

